Improved Bayesian Ethnorace Prediction

### 2021-02-07 paper.Rmd
first re-read

Research on racial and ethnic disparities in the United States can be constrained by the lack of available individual-level data with these markers [@cascio2012; @kuk2020]. This is especially true when racial geography is a key aspect of the research design. In these scenarios, researchers are often forced to use aggregate-level data on group proportions, such as county statistics, to draw inferences. But this method is susceptible to the modifiable areal unit problem, whereby statistical bias is introduced due to arbitrary and unequal geographic unit sizes [@fotheringham1991]. Individual-level data with ethnoracial categories is also particularly useful for geographic regression discontinuity designs---an increasingly common methodology which typically relies on individually geocoded addresses [@keele2018; @velez2019; @cantoni2020].

To overcome some of these challenges, in this paper I describe a method for imputing individual ethnorace categories directly. This method uses Bayes' rule to make predictions by combining information from nationwide distributions of these categories over other characteristics, such as names, geographies, political party identification, among others. My implementation builds off existing methods which use a similar algorithm [@elliott2009; @imai2016; @voicu2018]. The method described in Imai and Khanna (2016) in particular, and the associated `R` package `wru`, has become popular in recent studies on race and ethnicity. Its application has been used in work on racial protests and voting patterns [@enos2019], disparities in campaign financing [@grumbach2020; @grumbach2020a], and public health issues such as suicide rates [@studdert2020].

@@

In the next section I will provide some background on the inputs and outputs of my prediction algorithm. Then I will explain the methodology and compare my implementation with previous versions. Finally, I will demonstrate the predictive performance of my method when validated against the combined North Carolina and Florida voter file ($n$ = 21,000,000).

@@

Before discussing the method further, it is important to first define what "predicting ethnicity or race" means. These are categories which, although relatively immutable relative to other identities, do not have universally accepted delineations and meanings [@omi2014]. I follow the conventions of previous prediction methods by using the US Census Bureau categorizations [@elliott2009; @imai2016; @voicu2018]. In this framework, individuals can be classified as non-Hispanic White, non-Hispanic Black or African American, non-Hispanic Asian and Pacific Islander, non-Hispanic American Indian and Alaska Native, Hispanic or Latino alone, and non-Hispanic Other Race.[^2] Because Hispanic identity is defined by the Census, and understood commonly, as an ethnicity, rather than a race, I use the term "ethnorace" in this article to refer to any of the previously-mentioned categories.

@@

There are a few benefits to using the Census ethnoracial categorization. These groups capture a common understanding of racial and ethnic groups in the US, and correspond to the groups studied most frequently in social science research. Therefore we are also able to assemble the data sources of these groups' distributions that serve as inputs to the algorithm. This also facilitates comparison of my method against previous ethnorace prediction methods.[^3] One downside to using the Census categories, however, is that it obscures substantial heterogeneity that may exist within each group. Within Asian Americans and Latinos, for example, there is considerable variation in terms of national ancestry. Furthermore, the unfortunate necessity of an Other Race category ensures that important sources of diversity are washed over.[^4]

@@

The first names list I use comes from Tzioumis (2017). It is drawn from mortgage applicants and contains ethnorace counts in each of the six categories across 4,250 first names. Unlike Census data, which form the basis for much of my other data sources, this list of first names may be unrepresentative of the larger US population. To the extent that first name distributions differ by ethnorace given employment status, for example, this may be a concern. But the predictive benefits from using first name data, as I will demonstrate, likely overwhelms these concerns in most contexts.

@@

### Geolocation

My ethnorace distributions by geographies come from the 2010 decennial Census, accessed via IPUMS NHGIS.[^6] In descending order of average population, these geographies include *state*, *county*, *Census place*, *ZIP code*, and *Census block*. Predictions tend to improve with more precise levels of geography. With this in mind, my implementation automatically matches each individual to the most fine-grain level of geography available.

@@

# Method

>>>>

Research on racial and ethnic disparities in the United States can be constrained by the lack of available individual-level data with these markers [@cascio2012; @kuk2020]. This is particularly true when racial geography is a key aspect of the research design. In these contexts, researchers are often forced to use aggregate-level data on group proportions, such as county statistics, to draw inferences. But this method is susceptible to the modifiable areal unit problem, whereby statistical bias is introduced due to arbitrary and unequal geographic unit sizes [@fotheringham1991]. Furthermore, some research designs all but require individual-level data. Geographic regression discontinuity designs---an increasingly common causal inference technique---typically relies on individually geocoded addresses [@keele2018; @velez2019; @cantoni2020]. Therefore if race or ethnicity is a central element of the research question, individual-level identifiers for these categories are likely necessary.

To overcome some of these challenges, in this paper I describe a method for imputing individual ethnorace categories directly. This method uses Bayes' rule to make predictions by combining information from nationwide distributions of six ethnoracial categories over other characteristics, such as names, geographies, political party identification, among others. My implementation builds off existing methods which use a similar formula [@elliott2009; @imai2016; @voicu2018]. The method described in Imai and Khanna (2016) in particular, and the associated `R` package `wru`, has become popular in recent studies on race and ethnicity. Its application has been used in work on racial protests and voting patterns [@enos2019], disparities in campaign financing [@grumbach2020; @grumbach2020a], and public health issues such as suicide rates [@studdert2020].

@@

In the next section I will provide some background on the inputs and outputs of my prediction method. Then I will explain the methodology and compare my implementation with previous versions. Finally, I will demonstrate the predictive performance of my method when validated against the combined North Carolina and Florida voter file ($n$ = 21,000,000).

@@

Before discussing the methodology further, it is important to first define what "predicting ethnicity or race" means. These are categories which, although relatively immutable compared to other identities, do not have universally accepted delineations and meanings [@omi2014]. I follow the convention from previous ethnorace prediction methods by using the US Census Bureau categorizations [@elliott2009; @imai2016; @voicu2018]. In this framework, individuals can be classified as non-Hispanic White, non-Hispanic Black or African American, non-Hispanic Asian and Pacific Islander, non-Hispanic American Indian and Alaska Native, Hispanic or Latino alone, and non-Hispanic Other Race.[^2] Because Hispanic identity is defined by the Census, and understood commonly, as an ethnicity, rather than a race, I use the term "ethnorace" in this paper to refer to any of the previously-mentioned categories.

@@

There are a few benefits to using the Census ethnoracial categorization. These groups capture a common understanding of race and ethnicity in the US, and correspond to the groups studied most frequently in social science research. The data sources of these groups' distributions that serve as inputs to the prediction formula rely on the Census categorization. This also facilitates comparison of my method against previous ethnorace prediction methods.[^3] One downside to using the Census categories, however, is that it obscures substantial heterogeneity that may exist within each group. Within Asian Americans and Latinos, for example, there is considerable variation in terms of national ancestry. Furthermore, the unfortunate necessity of an Other Race category ensures that important sources of diversity are washed over.[^4]

@@

The first names list I use comes from Tzioumis (2017). It is drawn from mortgage applicants and contains ethnorace counts in each of the six categories across 4,250 first names. Unlike Census data, which form the basis for much of my other data sources, this list of first names may be unrepresentative of the larger US population. To the extent that first name distributions differ by ethnorace given employment status, for example, this may be a concern. But the predictive benefits from using first name data, as I will demonstrate, likely overwhelms these worries in most contexts.

@@

### Geolocations

My ethnorace distributions by geographies come from the 2010 decennial Census, accessed via IPUMS NHGIS.[^6] In descending order of mean population, these geographies include *state*, *county*, *Census place*, *ZIP code*, and *Census block*. Predictions tend to improve with more precise levels of geography. With this in mind, my implementation automatically matches each individual to the most fine-grain level of geography available. It is worth pointing out that this phenomenon is, in part, the consequence of generations of segregationist housing policies in the US. The fact that knowing an individual's ZIP code or Census block gives us so much knowledge about their race is an indictment of the US system more generally.

@@

### Party Identification

My party identification data come from a 2012 Gallup poll.[^7] The three categories of political party I include are Republican, Democrat, and Other (including independents and "don't knows").

[^7]: <https://news.gallup.com/poll/160373/democrats-racially-diverse-republicans-mostly-white.aspx>

### Age and Gender

Like my geolocation data, age and gender distributions come from the 2010 decennial Census, accessed via IPUMS NHGIS. These variables do not contain much predictive power in terms of ethnoracial classification, but nevertheless, I find that their inclusion in the formula helps slightly.

### Multi-unit Address

These data refer to ethnorace distributions over multi-unit housing occupancy. Individuals are matched to these probabilities if their address contains "Apt", "Unit", "#", or other such identifier. Unfortunately, I was not able to find these distributions for the 2010 decennial Census, so instead I use those from the year 2000.

# Methodology

$$
Pr(R|X) = \frac{Pr(X|R)Pr(R)}{Pr(X)}
$$
